Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-08-04 paper.tex
Add species annotations.

The Sequentially Markovian Coalescent (SMC) is an approximation of the
coalescent with recombination~\citep{mcvean2005approximating,marjoram2006fast},
and was primarily motivated by the need to
simulate longer genomes than was possible using tools like \ms.
The SMC is a good approximation to the
coalescent with recombination when we have four or fewer sampled
genomes~\citep{hobolth2014markovian,wilton2015smc}, but the
effects of the approximation are less well understood for larger
sample sizes, and several approaches have been proposed
that allow simulations to more closely approximate the coalescent
with recombination~\citep{chen2009fast,wang2014new,staab2015scrm}.
The SMC and SMC' models are available
in \msprime\ 1.0. However, they are currently implemented using a
naive rejection sampling approach, and are somewhat slower
to simulate than the exact coalescent with recombination. These
models are therefore currently only appropriate for studying the
SMC approximations themselves, although we intend to
implement them more efficiently in future versions.

\begin{figure}
\begin{center}
    \includegraphics{figures/ancestry-perf}
\end{center}
\caption{\label{fig-ancestry-perf}
    Approximate run time for \msprime\ simulations,
    for (A) small and (B) larger simulations.
    Each point is the runtime of one simulation,
    for various values of effective population size ($N$),
    genome length ($L$), and number of samples.
    Runtime scales quadratically with the product of $N$ and $L$;
    since recombination rate in these simulations was $10^{-8}$,
    the scaled recombination rate here is $\rho = NL/10^8$,
    shown on the horizontal axis.
    The black lines are quadratic fits separately in each panel
    and sample size.
}
\end{figure}

In human-like parameter regimes and for large sample sizes,
\msprime's implementation of the
exact coalescent with recombination comprehensively outperforms
all other simulators, including those based on SMC
approximations~\citep{kelleher2016efficient}. However, it is important
to note that although the implementation of Hudson's algorithm is
very efficient, it is still quadratic in the scaled recombination rate
$\rho = 4 N_e L$, where $L$ is the length of the genome in units of recombination distance.
This is because Hudson's algorithms tracks recombinations not only
in segments ancestral to the sample, but also between ancestral segments.
In other words, lineages may ``coalesce'' even if the pieces of ancestral material they carry
do not overlap---representing an ancestral chromosome
that is a genetic ancestor to each of two linages,
but not the most recent common genetic ancestor at any location along the genome.
When this happens, the resulting merged lineage carries ``trapped'' genetic material
that is not ancestral to any samples;
recombinations in the trapped material will cause the two segments of ancestry
to split apart into separate lineages again.
For large $\rho$, recombination events
in this trapped material will dominate.
\citet[Eq.~5.10]{hein2004gene} gave 
$\rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2$
as an upper bound
on the number of recombination events within trapped ancestral material
in Hudson's algorithm.
Since the sum is well approximated by $\log{n}$,
a natural conjecture is that the time complexity of Hudson's
algorithm is $O(\rho^2 \log^2 n)$.
This turns out to have the correct dependence on $\rho$---although
caution is needed; see below---but the dependence on $n$ is less accurate.

Figure~\ref{fig-ancestry-perf} gives a rough idea of the runtime of
a given simulation of a constant size population,
and verifies that runtime scales quadratically with $\rho$,
and depends much less strongly on the number of samples.
Taking a typical chromosome to be 1 Morgan in length,
these plot show, roughly, that simulating chromosome-length
samples from a population of thousands of individuals takes seconds,
while samples from a population of tens of thousands takes minutes.
However, it is important to note that the quadratic curves in the two
panels are different---predicting the runtimes of days-long simulations
using the timing of minutes-long runs is unlikely to be very accurate.
Further experiments show that adding a $\log^2 n$ factor
to predict runtimes across different sample sizes does not do well
except possibly at very large values of $\rho$.

>>>>

The Sequentially Markovian Coalescent (SMC) is an approximation of the
coalescent with recombination~\citep{mcvean2005approximating,marjoram2006fast},
and was primarily motivated by the need to
simulate longer genomes than was possible using tools like \ms.
The SMC is a good approximation to the
coalescent with recombination when we have fewer than five sampled
genomes~\citep{hobolth2014markovian,wilton2015smc}, but the
effects of the approximation are less well understood for larger
sample sizes, and several approaches have been proposed
that allow simulations to more closely approximate the coalescent
with recombination~\citep{chen2009fast,wang2014new,staab2015scrm}.
The SMC and SMC' models are available
in \msprime\ 1.0. However, they are currently implemented using a
naive rejection sampling approach, and are somewhat slower
to simulate than the exact coalescent with recombination. These
models are therefore currently only appropriate for studying the
SMC approximations themselves, although we intend to
implement them more efficiently in future versions.

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{figures/ancestry-perf}
\end{center}
\caption{\label{fig-ancestry-perf}
    Running time for \msprime\
    for (A) small and (B) larger simulations on an Intel i7-6600U CPU.
    Each point is the runtime of one simulation,
    for various values of effective population size ($N_e$),
    genome length in Morgans ($L$), and number of samples ($n$).
    Runtime scales quadratically with the product of $N_e$ and $L$,
    shown on the horizontal axis---for instance, figure A shows that
    1,000 samples of 1 Morgan-length chromosomes
    from a population of $N_e=2,000$ diploids
    would take about 2 seconds, and that the same number of
    0.01 Morgan segments in a population of size 200,000
    would take about the same time.
    Since recombination rate in these simulations was $10^{-8}$,
    $L$ is the number of base pairs divided by $10^8$.
    The black lines are quadratic fits separately in each panel
    and sample size. Also shown are the approximate values of
    $\rho$ for chromosome 1 in three species, using values from
    the \stdpopsim\ catalog~\citep{adrion2020community}.
}
\end{figure}

% Output from  python3 evaluation/plot.py ancestry-perf
% Predicted time for DroMel chr2L with n = 1000 = 107.85106995938845 hours
% Predicted time for DroMel chr2L with n = 100000 = -511.74273443828514 hours
% Predicted time for DroMel chr2L with n = 1000 = 177.40184421122058 hours
% Predicted time for DroMel chr2L with n = 100000 = 361.90415260982536 hours

In human-like parameter regimes and for large sample sizes,
\msprime's implementation of the
exact coalescent with recombination comprehensively outperforms
all other simulators, including those based on SMC
approximations~\citep{kelleher2016efficient}. However, it is important
to note that although the implementation of Hudson's algorithm is
very efficient, it is still quadratic in the scaled recombination rate
$\rho = 4 N_e L$, where $L$ is the length of the genome in units of recombination distance.
This is because Hudson's algorithms tracks recombinations not only
in segments ancestral to the sample, but also between ancestral segments.
In other words, lineages may ``coalesce'' even if the pieces of ancestral material they carry
do not overlap---representing an ancestral chromosome
that is a genetic ancestor to each of two linages,
but not the most recent common genetic ancestor at any location along the genome.
When this happens, the resulting merged lineage carries ``trapped'' genetic material
that is not ancestral to any samples;
recombinations in the trapped material will cause the two segments of ancestry
to split apart into separate lineages again.
For large $\rho$, recombination events
in this trapped material will dominate.
\citet[Eq.~5.10]{hein2004gene} gave
$\rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2$
as an upper bound
on the number of recombination events within trapped ancestral material
in Hudson's algorithm with $n$ samples.
Since the sum is well approximated by $\log{n}$,
a natural conjecture is that the time complexity of Hudson's
algorithm is $O(\rho^2 \log^2 n)$.
This turns out to have the correct dependence on $\rho$---although
caution is needed; see below---but the dependence on $n$ is less accurate.

Figure~\ref{fig-ancestry-perf} gives a rough idea of the runtime of
a given simulation of a constant size population,
and verifies that runtime scales quadratically with $\rho$,
and depends much less strongly on the number of samples.
Taking a typical chromosome to be 1 Morgan in length,
these plots show, roughly, that simulating chromosome-length
samples from a population of thousands of individuals takes seconds,
while samples from a population of tens of thousands takes minutes.
However, it is important to note that the quadratic curves in the two
panels are different---predicting the runtimes of days-long simulations
using the timing of minutes-long runs is unlikely to be very accurate.
Further experiments show that adding a $\log^2 n$ factor
to predict runtimes across different sample sizes does not do well
except possibly at very large values of $\rho$.
